Deciding Bitvector Termination with Program Synthesis

### 2014-07-06 papers/termination/paper.tex
Experiments table key
relateg work + figure 1

As Ramsey-based approaches are limited by the state of the art in safety
checking, recent research shifts back to more complex termination arguments
that are easier to
check~\cite{DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/HeizmannHP14}. 
Following the same trend, we investigate its extreme: \emph{unrestricted}
termination arguments.  This means that our ranking functions may involve
non-linearity and lexicographic orders: we do not commit to any particular
syntactic form, and do not use templates.

@@

\begin{figure*}
\centering
 \begin{tabular}{|ll||c|c|c|c|c|c|c|c|}
 \hline
  & & \multicolumn{8}{c|}{Program} \\
  & & \multicolumn{2}{c|}{Rationals/Integers} & \multicolumn{2}{c|}{Reals} & \multicolumn{2}{c|}{Bit-vectors} & \multicolumn{2}{c|}{Floats} \\
  & & L & NL & L & NL & L & NL & L & NL \\
  \hline
  \hline
  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} && & &\checkmark&\checkmark&\checkmark&\checkmark\\
   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & && \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
   & Non-linear lexicographic &  &  & &&\checkmark&\checkmark&\checkmark&\checkmark\\
   & Non-linear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} &  \cite{DBLP:conf/vmcai/BradleyMS05} & &&\checkmark&\checkmark&\checkmark&\checkmark\\
   \hline
 \end{tabular}

 \caption{Legend: \checkmark = we can handle\label{fig:handletable}}
\end{figure*}

@@

Somehow expected, most of the techniques are specialised in the synthesis of linear ranking functions for linear programs 
over integers (or rationals) \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,
DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13}. Among them, 
Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures 
%to compute transition invariants as proofs for the termination of linear programs 
\cite{DBLP:conf/cav/LeeWY12},
Leike and Heizmann propose linear ranking templates 
%in order to synthesise
%termination arguments for linear loop programs 
\cite{DBLP:conf/tacas/LeikeH14}, whereas
Bradley et al. compute lexicographic linear ranking functions supported by inductive linear invariants 
%for loops with linear guards and transitions: 
\cite{DBLP:conf/cav/BradleyMS05}. 
%Learning: \cite{DBLP:conf/cav/HeizmannHP14}

@@

Given the lack of research in handling \emph{not-linear programs}, as well
as \emph{programs over bit vectors and floats}, our work focuses on covering
these areas.  One of the obvious conclusions that can be reached from
observing Figure~\ref{fig:handletable}, is that most of the works tend to
specialise on a certain aspect of termination proving that they can solve
efficiently.  Conversely to this view, we aim for \emph{generality}, as we
do not restrict the form of the synthesised ranking functions, nor the form
of the input programs.

@@

Proving program termination implies the simultaneous search for a
termination argument and a supporting invariant.  For this purpose, several
tools alternate between calls to a safety prover and calls to a ranking
function synthesis tool \cite{}.  In order to share more information about
the state of the termination proofs between the two tools, Brockschmidt et
al.  use a single representation of the state of the termination proof
search that both tools operate over, resulting in performance improvements
\cite{DBLP:conf/cav/BrockschmidtCF13}.  In \cite{DBLP:conf/cav/BradleyMS05},
Bradley et al.  combine the generation of ranking functions with the
generation of invariants to form one constraint solving problem such that
the necessary supporting invariants for the ranking function are discovered
on demand.  In our setting, both the ranking function and the supporting
invariant are iteratively constructed through the same refinement loop.

While program termination has been extensively studied, much less research has been conducted in the area of proving nontermination.
Gupta et al. dynamically enumerate lasso-shaped candidate
paths for counterexamples, and then statically prove their feasibility \cite{DBLP:conf/popl/GuptaHMRX08}. 
In \cite{DBLP:conf/tacas/ChenCFNO14}, Chen et al. prove nontermination via
reduction to safety proving. Their iterative algorithm uses counterexamples to a fixed safety property
to refine an underapproximation of a program.

>>>>

As Ramsey-based approaches are limited by the state of the art in safety
checking, recent research shifts back to more complex termination arguments
that are easier to
check~\cite{DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/CookSZ13}. 
Following the same trend, we investigate its extreme: \emph{unrestricted}
termination arguments.  This means that our ranking functions may involve
non-linearity and lexicographic orders: we do not commit to any particular
syntactic form, and do not use templates.

@@

\begin{figure*}
\centering
 \begin{tabular}{|ll||c|c|c|c|c|c|c|c|}
 \hline
  & & \multicolumn{8}{c|}{Program} \\
  & & \multicolumn{2}{c|}{Rationals/Integers} & \multicolumn{2}{c|}{Reals} & \multicolumn{2}{c|}{Bit-vectors} & \multicolumn{2}{c|}{Floats} \\
  & & L & NL & L & NL & L & NL & L & NL \\
  \hline
  \hline
  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} & - & & &\checkmark&\checkmark&\checkmark&\checkmark\\
   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & && \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
   & Non-linear lexicographic & - & - & &&\checkmark&\checkmark&\checkmark&\checkmark\\
   & Non-linear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} &  \cite{DBLP:conf/vmcai/BradleyMS05} & &&\checkmark&\checkmark&\checkmark&\checkmark\\
   \hline
 \end{tabular}

 \caption{Summary of related termination analyses. Legend: \checkmark = we can handle; - = no available works.} \label{fig:handletable}
\end{figure*}

@@

Somehow expected, most of the techniques are specialised in the synthesis of linear ranking functions for linear programs 
over integers (or rationals) \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,
DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}. Among them, 
Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures 
%to compute transition invariants as proofs for the termination of linear programs 
\cite{DBLP:conf/cav/LeeWY12},
Leike and Heizmann propose linear ranking templates 
%in order to synthesise
%termination arguments for linear loop programs 
\cite{DBLP:conf/tacas/LeikeH14}, whereas
Bradley et al. compute lexicographic linear ranking functions supported by inductive linear invariants 
%for loops with linear guards and transitions: 
\cite{DBLP:conf/cav/BradleyMS05}. 
%Learning: \cite{DBLP:conf/cav/HeizmannHP14}

@@

Given the lack of research in handling \emph{not-linear programs}, as well
as \emph{programs over bit vectors and floats}, our work focused on covering
these areas.  One of the obvious conclusions that can be reached from
observing Figure~\ref{fig:handletable}, is that most of the works tend to
specialise on a certain aspect of termination proving that they can solve
efficiently.  Conversely to this view, we aim for \emph{generality}, as we
do not restrict the form of the synthesised ranking functions, nor the form
of the input programs.

@@

Proving program termination implies the simultaneous search for a
termination argument and a supporting invariant.  
%For this purpose, several tools alternate between calls to a safety prover and calls to a ranking function synthesis tool \cite{}.  
%In order to share more information about
%the state of the termination proofs between the two tools, 
Brockschmidt et al. 
share the same representation of the state of the termination proof
between the safety prover and the ranking function synthesis tool \cite{DBLP:conf/cav/BrockschmidtCF13}.  
In \cite{DBLP:conf/cav/BradleyMS05},
Bradley et al.  combine the generation of ranking functions with the
generation of invariants to form one constraint solving problem such that
the necessary supporting invariants for the ranking function are discovered
on demand.  In our setting, both the ranking function and the supporting
invariant are iteratively constructed through the same refinement loop.

While program termination has been extensively studied, much less research has been conducted 
in the area of proving non-termination. Gupta et al. dynamically enumerate lasso-shaped candidate
paths for counterexamples, and then statically prove their feasibility \cite{DBLP:conf/popl/GuptaHMRX08}. 
In \cite{DBLP:conf/tacas/ChenCFNO14}, Chen et al. prove nontermination via
reduction to safety proving. Their iterative algorithm uses counterexamples to a fixed safety property
to refine an underapproximation of a program.
